Revealing transient strain in geodetic data with Gaussian process regression

### 2017-10-10 RevisedSubmission/hines_hetland_2017_gji.tex
did more

\section{Outlier detection algorithm}
Our outlier detection algorithm is loosely based on the data editing algorithm from \citet{Acheson1975}. Let $\mitbf{d}^*$ denote all $n$ GNSS displacement observations for a single directional component, which have been made at positions and times $\mitbf{P}$.  We model $\mitbf{d}^*$ as a realization of the random vector 
\begin{equation}\label{eq:OutlierDataVec}
\mitbf{d} = \mitbf{G}\mitbf{m} + v(\mitbf{P}) + \mitbf{w},
\end{equation}
where $\mitbf{G}$ and $\mitbf{m}$ are the same as in eq. (\ref{eq:DataVec}), $v$ is a Gaussian process distributed as $\mathcal{GP}(0,C_v)$, and $\mitbf{w}$ is a vector of uncorrelated Gaussian noise with known standard deviations $\mitbf{\sigma} = [\sigma_1, \sigma_2, ..., \sigma_n]$. The Gaussian process $v$ is intended to describe transient features in the data that cannot be explained by the linear trend or seasonal terms in $\mitbf{G}$. We let the temporal covariance of $v$ be a squared exponential, and we let $v$ be spatially uncorrelated so that,
\begin{equation}
C_v(p,p') = \phi^2\exp\left(\frac{-|t - t'|^2}{2\tau^2}\right) C_w(\vec{x},\vec{x}'),
\end{equation}
where 
\begin{equation}\label{eq:WhiteCov}
C_w(\vec{x},\vec{x}') = 
\begin{cases}
1, &\vec{x} = \vec{x}' \\
0, &\vec{x} \neq \vec{x}'
\end{cases}.
\end{equation}
The spatial covariance of $v$ has little effect on the detected outliers, and so we have assumed that $v$ is spatially uncorrelated for simplicity. Based on our experience, $v$ can reasonably describe most transient features in the data when we set $\phi = 1$ mm and $\tau = 10$ days.

Our goal is to find the index set of non-outliers in $\mitbf{d}^*$, which we denote as $\mitbf{\Omega}$. We use a tilde to indicate that an array only contains elements corresponding to $\mitbf{\Omega}$ (e.g., the vector of non-outlier observations is denoted $\tilde{\mitbf{d}}^* = [d_i^*]_{i \in \mitbf{\Omega}}$). The outliers are identified iteratively, and we initiate $\mitbf{\Omega}$ with all $n$ indices. We consider outliers to be data that are inconsistent with eq. (\ref{eq:OutlierDataVec}), which we determine with the residual vector 
\begin{align}\label{eq:Residual}
\mitbf{r} &= \mitbf{d}^* - \mathrm{E}\left[\Big(\mitbf{G}\mitbf{m} + v(\mitbf{P})\Big) \Big| \Big(\tilde{\mitbf{d}} = \tilde{\mitbf{d}}^*\Big)\right] \\
          &= \mitbf{d}^*  - 
            \left[\begin{array}{cc}
                  C_v(\mitbf{P},\tilde{\mitbf{P}}) & \mitbf{G} \nonumber \\
                  \end{array}\right]
            \left[\begin{array}{cc}
                  C_v(\tilde{\mitbf{P}},\tilde{\mitbf{P}}) + \mathrm{diag}(\tilde{\mitbf{\sigma}}^2) & \tilde{\mitbf{G}} \\
                  \tilde{\mitbf{G}}^T  & \mitbf{0} \\
                  \end{array}\right]^{-1}
            \left[\begin{array}{c}
                  \tilde{\mitbf{d}}^* \\
                  \mitbf{0} \\
                  \end{array}\right].
\end{align}
Data with abnormally large residuals are identified as outliers. For each iteration, we compute $\mitbf{r}$ and then update $\mitbf{\Omega}$ so that it contains the indices of $\mitbf{r}$ whose weighted values are less than $\lambda$ times the weighted root mean square of $\tilde{\mitbf{r}}$, 
\begin{equation}\label{eq:Update}
\mitbf{\Omega} \leftarrow \left\{i : \left|\frac{r_i}{\sigma_i}\right| < \lambda \cdot \sqrt{\frac{1}{|\mitbf{\Omega}|} \sum_{j \in \mitbf{\Omega}} \frac{r_j^2}{\sigma_j^2}} \right\}.
\end{equation} 
Iterations continue until the new $\mitbf{\Omega}$ is the same as the previous $\mitbf{\Omega}$. We use $\lambda=4$ in this study, which in our experience accurately identifies outliers without unnecessarily decimating the data.

>>>>

\section{Outlier detection algorithm}
Our outlier detection algorithm is loosely based on the data editing algorithm from \citet{Acheson1975}. Let $\mitbf{d}^*$ denote all $n$ GNSS displacement observations for a single directional component, which have been made at positions and times $\mitbf{P}$.  We describe $\mitbf{d}^*$ as a realization of the random vector 
\begin{equation}\label{eq:OutlierDataVec}
\mitbf{d} = \mitbf{G}\mitbf{m} + v(\mitbf{P}) + \mitbf{w},
\end{equation}
where $\mitbf{G}$ and $\mitbf{m}$ are the same as in eq. (\ref{eq:DataVec}), $v$ is a Gaussian process distributed as $\mathcal{GP}(0,C_v)$, and $\mitbf{w}$ is a vector of uncorrelated Gaussian noise with known standard deviations $\mitbf{\sigma} = [\sigma_1, \sigma_2, ..., \sigma_n]$. The Gaussian process $v$ is intended to describe transient features in the data that cannot be explained by the linear trend or seasonal terms in $\mitbf{G}$. We let the temporal covariance of $v$ be a squared exponential, and we let $v$ be spatially uncorrelated so that,
\begin{equation}
C_v(p,p') = \phi^2\exp\left(\frac{-|t - t'|^2}{2\tau^2}\right) C_w(\vec{x},\vec{x}'),
\end{equation}
where 
\begin{equation}\label{eq:WhiteCov}
C_w(\vec{x},\vec{x}') = 
\begin{cases}
1, &\vec{x} = \vec{x}' \\
0, &\vec{x} \neq \vec{x}'
\end{cases}.
\end{equation}
The spatial covariance of $v$ has little effect on the detected outliers, and so we have assumed that $v$ is spatially uncorrelated for simplicity. Based on our experience, $v$ can reasonably describe most transient features in the data when we set $\phi = 1$ mm and $\tau = 10$ days.

Our goal is to find the index set of non-outliers in $\mitbf{d}^*$, which we denote as $\mitbf{\Omega}$. We use a tilde to indicate that an array only contains elements corresponding to $\mitbf{\Omega}$ (e.g., the vector of non-outlier observations is denoted $\tilde{\mitbf{d}}^* = [d_i^*]_{i \in \mitbf{\Omega}}$). The outliers are identified iteratively, and we initiate $\mitbf{\Omega}$ with all $n$ indices. We consider outliers to be data that are poorly explained by the model $\mitbf{G}\mitbf{m} + v(\mitbf{P})$, which is determine by the residual vector 
\begin{align}\label{eq:Residual}
\mitbf{r} &= \mitbf{d}^* - \mathrm{E}\left[\Big(\mitbf{G}\mitbf{m} + v(\mitbf{P})\Big) \Big| \Big(\tilde{\mitbf{d}} = \tilde{\mitbf{d}}^*\Big)\right] \\
          &= \mitbf{d}^*  - 
            \left[\begin{array}{cc}
                  C_v(\mitbf{P},\tilde{\mitbf{P}}) & \mitbf{G} \nonumber \\
                  \end{array}\right]
            \left[\begin{array}{cc}
                  C_v(\tilde{\mitbf{P}},\tilde{\mitbf{P}}) + \mathrm{diag}(\tilde{\mitbf{\sigma}}^2) & \tilde{\mitbf{G}} \\
                  \tilde{\mitbf{G}}^T  & \mitbf{0} \\
                  \end{array}\right]^{-1}
            \left[\begin{array}{c}
                  \tilde{\mitbf{d}}^* \\
                  \mitbf{0} \\
                  \end{array}\right].
\end{align}
Data with abnormally large residuals are identified as outliers. For each iteration, we compute $\mitbf{r}$ and then update $\mitbf{\Omega}$ so that it contains the indices of $\mitbf{r}$ whose weighted values are less than $\lambda$ times the weighted root mean square of $\tilde{\mitbf{r}}$, 
\begin{equation}\label{eq:Update}
\mitbf{\Omega} \leftarrow \left\{i : \left|\frac{r_i}{\sigma_i}\right| < \lambda \cdot \sqrt{\frac{1}{|\mitbf{\Omega}|} \sum_{j \in \mitbf{\Omega}} \frac{r_j^2}{\sigma_j^2}} \right\}.
\end{equation} 
Iterations continue until the new $\mitbf{\Omega}$ is the same as the previous $\mitbf{\Omega}$. We use $\lambda=4$ in this study, which in our experience accurately identifies outliers without unnecessarily decimating the data.
